---
title: "Practical 1 - Medical Databases: Predicting Sepsis from ICU Data"
author: "Mallory Murphy"
date: "2026-06-04"
output: html_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110,
  cache      = TRUE,
  cache.lazy = FALSE
)

GLOBAL_SEED <- 42
set.seed(GLOBAL_SEED)

options(
  digits     = 4,
  scipen     = 999,
  dplyr.summarise.inform = FALSE
)
```

> **Submission.** Knit your Rmd file into HTML and submit it via Brightspace. Include a link to your git repository (GitHub or git.cs.dal.ca) as a comment on your submission.

---

## 0 Setup

```{r packages}
library(tidyverse)    # dplyr, ggplot2, tidyr, purrr, ...
library(tidymodels)   # rsample, recipes, parsnip, workflows, tune, yardstick, ...
library(ranger)       # random forest engine
library(xgboost)      # gradient-boosted trees engine
library(glmnet)       # penalised logistic regression engine
library(here)
library(fs)

theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank()))
```

---

## 1 Introduction

[Sepsis](https://www.who.int/news-room/fact-sheets/detail/sepsis) is life-threatening organ dysfunction caused by a dysregulated immune response to infection.
It accounts for roughly 20 % of global deaths, and each hour of delayed treatment increases mortality by 4–8 %.

The [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) dataset contains hourly ICU records for ~40 000 patients from two US teaching hospitals.
Each record (row) has 8 vital signs, 26 laboratory measurements, 6 demographic fields, and a binary `SepsisLabel` (set to 1 up to 6 hours *before* clinical onset so that early-warning models can be trained).
We simplify the temporal problem to a static one by taking one random hourly record per patient, giving a standard tabular classification task with realistic clinical challenges: class imbalance, high missingness, and mixed data types.

**Q1. (2 marks)** Look up the [PhysioNet variable descriptions](https://physionet.org/content/challenge-2019/1.0.0/) and include an explanation of the dataset into this notebook (one row per variable: name, description, units).  

**Vital signs (columns 1-8)**
HR:	Heart rate (beats per minute)
O2Sat:	Pulse oximetry (%)
Temp:	Temperature (Deg C)
SBP:	Systolic BP (mm Hg)
MAP:	Mean arterial pressure (mm Hg)
DBP:	Diastolic BP (mm Hg)
Resp:	Respiration rate (breaths per minute)
EtCO2:	End tidal carbon dioxide (mm Hg)
**Laboratory values (columns 9-34)**
BaseExcess:	Measure of excess bicarbonate (mmol/L)
HCO3:	Bicarbonate (mmol/L)
FiO2:	Fraction of inspired oxygen (%)
pH:	N/A
PaCO2:	Partial pressure of carbon dioxide from arterial blood (mm Hg)
SaO2:	Oxygen saturation from arterial blood (%)
AST:	Aspartate transaminase (IU/L)
BUN:	Blood urea nitrogen (mg/dL)
Alkalinephos:	Alkaline phosphatase (IU/L)
Calcium: Calcium level	(mg/dL)
Chloride: Chloride level	(mmol/L)
Creatinine: Creatine level	(mg/dL)
Bilirubin_direct:	Bilirubin direct (mg/dL)
Glucose:	Serum glucose (mg/dL)
Lactate:	Lactic acid (mg/dL)
Magnesium: Magnesium level	(mmol/dL)
Phosphate: Phosphate level	(mg/dL)
Potassium: Potassium level	(mmol/L)
Bilirubin_total:	Total bilirubin (mg/dL)
TroponinI:	Troponin I (ng/mL)
Hct:	Hematocrit (%)
Hgb:	Hemoglobin (g/dL)
PTT:	Partial thromboplastin time (seconds)
WBC:	Leukocyte count (count*10^3/µL)
Fibrinogen: Fibrinogen level	(mg/dL)
Platelets: Platelet count	(count*10^3/µL)
**Demographics (columns 35-40)**
Age: Age in	years (100 for patients 90 or above)
Gender:	Female (0); Male (1)
Unit1:	Administrative identifier for ICU unit (MICU)
Unit2:	Administrative identifier for ICU unit (SICU)
HospAdmTime:	Hours between hospital admit and ICU admit
ICULOS:	ICU length-of-stay (hours since ICU admit)
**Outcome (column 41)**
SepsisLabel: Sepsis patients, SepsisLabel is 1 if t≥tsepsis−6 and 0 if t<tsepsis−6; Non-sepsis patients, SepsisLabel is 0.

**Q2. (2 marks)** Then choose **three variables** you hypothesise might be strong predictors of sepsis and write one sentence each explaining your reasoning.  There is no right or wrong answer to this, but you should think about your and problem before diving into the EDA and prediction!
*(Hint: look up how sepsis is clinically diagnosed.)*  

Clinically diagnosing sepsis commonly involves checking vital signs like temperature, heart rate, blood pressure, etc. These vital signs would likely give some indication if the body is working to fight an infection. For this reason, the three variables I hypothesize to be predictors of sepsis is diastolic blood pressure (because it was found to be a slightly better predictors that systolic blood pressure in the Gao et al. article), heart rate, and temperature. 

---

## 2 Data Loading and Splitting

To avoid inadvertent data leakage from your test set, we are going to perform our test-train split before even our EDA and cleaning related analyses. This is not always done but is a good practice if you aren't using an external/independent test dataset.

### 2.1 Load the data

I've already done the random sampling of the dataset to remove the time-series elements but the original dataset can be acquired directly from [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) 
```{r data-paths}
data_dir  <- here("data")
cache_dir <- fs::path(data_dir, "cache")
fs::dir_create(cache_dir)

data_url <- "http://maguire-lab.github.io/health_data_science_research_2026/static_files/practicals/lab1_data/physionet_subsample_modified_1.0.0.rds"
rds_path <- fs::path(cache_dir, "physionet_subsample_1.0.0.rds")
```

Rmd lets you cache expensive or slow operations (like downloading big files or complicated processing) so they only re-run if the code and variables have changed.  This can be very useful although can sometimes cause some inadvertent issues that reduce reproducibility!  You should also do a completely clean re-run of everything once your analyses are finalised if possible.

```{r load-data, cache = TRUE}
if (!file.exists(rds_path)) {
  download.file(data_url, rds_path, mode = "wb", quiet = TRUE)
}
sepsis <- readRDS(rds_path)
```

### 2.2 Train / test split

We split the data **before** any exploration so that analysis choices cannot be influenced by the test set - this is a key discipline in applied machine learning.  Stratified splitting preserves the sepsis rate in both halves.

```{r split}
# will make later things easier to do!s
sepsis <- sepsis |>
  mutate(SepsisLabel = factor(SepsisLabel,
                              levels = c(1, 0),
                              labels = c("Sepsis", "No Sepsis")))

# splitting
split <- initial_split(sepsis, prop = 0.75, strata = SepsisLabel)
train <- training(split)
test  <- testing(split)

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test),  "\n")
```

**Q3. (1 mark)** What percentage of our dataset is being assigned to test?
```{r, percent of dataset is split test}

750/3000

```
25% of the dataset is assigned to test.
--- 

## 3 Exploratory Data Analysis

Again, do your EDA on `train` only - we do not look at `test` until Section 6.

### 3.1 Class balance


**Q4. (2 marks).** Uncomment and complete the code below to check the class balance in the **training** set.
`count()` tallies rows per group; add a `prop` column to show each class as a fraction of the total.

```{r q4-class-balance}
train |>
  count(SepsisLabel) |>
  mutate(prop = n / sum(n))

```


**Q5. (2 mark)** Using your solution to Q4, complete this plotting code to create a barplot of the relative sepsis label balance.  Change the colour of the sepsis bar to green.

```{r class-balance-plot, fig.height = 2.8}
train |>
  count(SepsisLabel) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(SepsisLabel, prop, fill = SepsisLabel)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.01)), vjust = -0.4) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
  scale_fill_manual(values = c("No Sepsis" = "grey60", 
                               "Sepsis"    = "green")) +
  labs(x = NULL, y = "Percentage of Trainining Data", title = "Sepsis Class Balance") +
  theme(legend.position = "none")
```

**Q6. (3 marks)**  Given the class imbalance, which of the following 2 metrics might be the most misleading and why? accuracy, precision, recall, F1-score, ROC-AUC, PR-AUC?

Accuracy ((TP+TN)/total N) would be misleading because there is such large class imbalace that if the model predicts all positive cases incorrectly but all negative cases correctly, the model would still achieve a high accuracy score (in this case, ~98%). 
https://towardsdatascience.com/precision-recall-curve-is-more-informative-than-roc-in-imbalanced-data-4c95250242f6/

ROC-AUC would also be misleading because it plots true positives against false positives and in this case where true negatives dominate the dataset, the demoninator of the false positive rate (FP/(FP+TN)) is over-inflated due to many more true negatives which essentially overshadows any detection of false positives. 
https://www.preprints.org/manuscript/202510.0958

### 3.2 Missingness

```{r missingness-bar, fig.height = 5}
train |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, fct_reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Missing data by variable (training set)",
       x = "Percent missing", y = NULL)
```

**Q7. (1 mark)**
Modify the above to plot only missingness for Sepsis positive data
```{r missingness-sepsis-bar, fig.height = 5}
train |>
  filter(SepsisLabel == "Sepsis") |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, fct_reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Missing data by variable among Sepsis+ (training set)",
       x = "Percent missing", y = NULL)
```

**Q8. (3 marks)** Which three variables have the highest missingness? In an ICU context, why might a lab value be missing - is this likely to be random, or informative? Does this mean you should drop or impute these values?

In the entire training set (Q6), the variables with the most missing values were: Troponinl, Bilirubin_direct, and Fibrinogen. Those were also the same variables among those with Sepsis, however, Fibrinogen was listed second instead of third for most missing.

A lab value could be missing for several reasons. Firslty, the test to get that measurement might not have been completed. There could have been an error with the test measuring that variable or an error in inputting the data into the EHR. If there are differences in missingness between people with vs without Sepsis, this could be informative. Missingness at random could still be informative for administrative purposes but less so for clinical purposes. In this case, missingness does not seem to differ by Sepsis status. Missing at random would be OK to be dropped and move forward with a complete case analysis as so long as missingness does not exceed >5%. However, which such a high degree of missingness (approximately 100%), it would be best to use multiple imputation instead of dropping/removing missing values.


### 3.3 Vital-sign distributions by outcome

```{r vitals-density, fig.height = 5}
vitals <- c("HR", "O2Sat", "Temp", "SBP", "MAP", "DBP", "Resp")

train |>
  select(all_of(vitals), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Vital-sign distributions by sepsis status")
```

**Q9 (2 marks).** Modify the code above to plot the distribution of **three other lab variables** of your choice. 

```{r q9-lab-density}
vitals2 <- c("Glucose", "Magnesium", "Phosphate")

train |>
  select(all_of(vitals2), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Vital-sign(2) distributions by sepsis status")
```

Correlation plot of all variables by similarity of their correlation in Sepsis and Non-Sepsis

```{r}
cor_mat <- function(df) {
  df |>
    select(where(is.numeric)) |>           
    cor(use = "pairwise.complete.obs")     
}

m_ns  <- cor_mat(train |> filter(SepsisLabel == "No Sepsis"))
m_sep <- cor_mat(train |> filter(SepsisLabel == "Sepsis"))
m_dif <- m_sep - m_ns

m_dif[is.na(m_dif)] <- 0
hc <- hclust(as.dist(1 - abs(m_dif)), method = "average")
var_order <- hc$labels[hc$order]

m_dif |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  tidyr::pivot_longer(-var1, names_to = "var2", values_to = "diff") |>
  mutate(var1 = factor(var1, levels = var_order),
         var2 = factor(var2, levels = var_order)) |>
  ggplot(aes(var1, var2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-0.5, 0.5),
                       oob = scales::squish, name = "Δ r") +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       title = "Change in pairwise correlations: Sepsis − No Sepsis") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid  = element_blank())
```

### 3.4 Dimensionality reduction

PCA compresses all numeric features into a small number of axes and lets us ask: *do septic and non-septic patients separate at all in the joint feature space?*
We build it inside a small `recipe` so the preprocessing (imputation, normalisation) is explicit and reproducible.

Recipes are a special type of function in `tidymodels` that differ from normal R functions in key ways (internal state).

A plain R function executes immediately and is stateless i.e., every call recomputes from scratch.
```
rscale_hr <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

train$HR_z <- scale_hr(train$HR)
test$HR_z  <- scale_hr(test$HR)   # BUG: uses test set's own mean/sd
```

The test set is standardised using statistics computed *from the test set*, which leaks information and means train and test are now on different scales. To do it correctly you'd have to manually save `mean(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and reuse them and then do the same for every other transformation, for every column, for imputation, encoding, etc. It gets unmanageable fast.

A recipe on the otherhand is a two-phase object: first you declare the steps, then you `prep()` it on training data (which estimates and stores any required parameters means, sds, imputation values, factor levels, PCA loadings…), then you `bake()` it on any new data to apply the same learned transformations.

```{r pca-prep}
pca_prep <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 10, id = "pca") |>
  prep()

pca_scores <- bake(pca_prep, new_data = NULL)
```

```{r pca-scatter, fig.width = 5.5, fig.height = 4}
pca_scores |>
  ggplot(aes(PC01, PC02, colour = SepsisLabel)) +
  geom_point(alpha = 0.75, size = 0.7) +
  scale_colour_manual(values = c("No Sepsis" = "grey60",
                                 "Sepsis"    = "firebrick")) +
  labs(title = "PCA: PC1 vs PC2", colour = NULL) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2)))
```

```{r pca-scree, fig.height = 3}
tidy(pca_prep, id = "pca", type = "variance") |>
  filter(terms == "percent variance") |>
  ggplot(aes(component, value)) +
  geom_col(fill = "steelblue") +
  labs(x = "Principal component", y = "% variance explained",
       title = "PCA scree plot")
```


The two classes overlap heavily in the PC1/PC2 projection — this is common in clinical data where no single linear combination cleanly separates cases.

**Q10 (2 marks).** What does this tell us about the potential decision boundary our predictive model would have to fit?  Which models easily support this type of boundary?

This tells us that the decision boundary our model would have to fit should not be linear. Examples of models that could easily support this type of boundary would be random forest, and/or neural network. 

---

## 4 Feature Engineering

Let's create a few new features likely to be informative for sepsis:

**Q11. (1 mark)** Modify the following code to also add an additional feature called `shock_index` which is defined as heart rate divided by the systolic blood pressure. 

```{r feature-engineering}
add_features <- function(df) {
  df |>
    mutate(
      pulse_pressure  = SBP - DBP,
      bun_creat_ratio = BUN / Creatinine,  # > 20 suggests pre-renal AKI
      age_group = cut(Age,
                      breaks = c(-1, 40, 65, 80, 120),
                      labels = c("<=40", "41-65", "66-80", "80+")),
      shock_index = HR / SBP
    )
}
train <- add_features(train)
test  <- add_features(test)
```

## 5 Modelling

Now lets try to fit some predictive models to the training set

### 5.1 Cross-validation folds

Five-fold stratified CV on the training set, used to select hyperparameters (see below) by repeatedly training the model with different hyperparameter values on 4/5ths of the data and evaluating on each of the held-out 1/5ths.s   

```{r cv-folds}
folds <- vfold_cv(train, v = 5, strata = SepsisLabel)
```

### 5.2 Preprocessing recipes

Not all models can handle the same type of data and some models (like linear models) benefit from normalising values to similar ranges. Most models will ignore missing values but we are going to add a missingness-indicator columns so the model can learn from *which* values are absent.

We are going to use `tidymodel` recipes again so we can apply the training normalisation etc to the test set to prevent leakage.

```{r recipes}
base_recipe <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_indicate_na(all_numeric_predictors(), prefix = "miss") |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors())

glm_recipe <- base_recipe |>
  step_normalize(all_numeric_predictors())
```

### 5.3 Model specifications

Now let's specify our 3 models:
- Logistic regression: a linear baseline. Outputs P(Sepsis∣X) via a logistic link function (i.e., squeezing our linear model through a sigmoid function that constrains between 0 and 1). By default `glmnet` uses a regularised model.  This means we are penalising the model complexity (the size of the weights/coefficients are added to the cost function).  Specifically, we are using an cross-validation optimied combination of L1 and L2 penalties (absolute and squared weight/coefficient sizes).  This is useful as many of our ~40 lab variables are correlated or weakly informative.

- Random forest: an ensemble of decision trees, each fit on a bootstrap sample with a random subset of features at each split. Captures non-linearities and interactions automatically, and handles mixed-scale features without preprocessing. Robust to outliers and (via `ranger`) to NAs.

- XGBoost: gradient-boosted trees. Builds trees sequentially, each one correcting the residual errors of the previous ensemble. Typically the strongest tabular performer, but the most sensitive to model (hyper)parameters and easy to overfit without tuning.

Each of these has several hyperparameters (parts of the model we don't directly learn from the data during fitting; instead we choose them before fitting and select the best values by cross-validation). Marking a hyperparameter with `tune()` is a placeholder that says "leave this blank for now — we'll fill it in during the tuning step." The values will later be chosen by tune_grid() based on cross-validated performance on the training folds.

**Q12 4 marks**: Using R's documentation (e.g., `?logistic_reg`) look up and briefly explain each of the hyperparamters we are tuning in cross-validation for each of the 3 models

?logistic_reg = glmnet = penalty and mixture
- penalty = involves two primary penalty parameters: lambda, which controls the overall penalty strength, and alpha, which controls the mixture between Ridge (L2) and Lasso (L1) penalties
- mixture = controls the balance between Lasso (L1) and Ridge (L2) penalties

?rand_forest = ranger = min_n
- min_n = minimum number of data points required to further split a node

?boost_tree = xgboost = trees, tree_depth, learn_rate, loss_reduction, sample_size, and min_n
- trees = number of trees
- tree_depth = maximum number of nodes from the root to leaves for each tree
- learn_rate = scales the contribution of each additional tree and controls for overfitting and model training speed
- loss_reduction = minimum reduction in the loss function required to split on a tree's node 
- sample_size = proportion of training data randomly sampled to build each tree
- min_n = minimum number of data points required to further split a node

```{r model-spec}
glm_spec <- logistic_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

# hyperparameters: penalty scaling factor and proportion L1 vs L2s
rf_spec <- rand_forest(trees = 500, min_n = tune()) |>
  set_engine("ranger", importance = "permutation", num.threads = 1) |>
  set_mode("classification")

xgb_spec <- boost_tree(
  trees = 500, tree_depth = tune(), learn_rate = tune(),
  loss_reduction = tune(), sample_size = tune(),
   min_n = tune()
) |>
  set_engine("xgboost", nthread = 1) |>
  set_mode("classification")
```


### 5.4 Tune all three models

Now let's specify our workflow to use cross-validation to tune all models and compare across them.

```{r workflow-set}
wf_set <- workflow_set(
  preproc = list(glm = glm_recipe, rf = base_recipe, xgb = base_recipe),
  models  = list(glm = glm_spec,   rf = rf_spec,      xgb = xgb_spec),
  cross   = FALSE
)
```
We are going to calculate a set of different performance metrics across CV folds.

- `pr_auc`: Area under the Precision-Recall curve. Sweeps the classification threshold from strict to lenient and plots precision against recall at each. Single-number summary of how well the model ranks septic patients above non-septic ones, without being inflated by the large true-negative pool. Baseline ≈ class prevalence (~0.02 here), so any value meaningfully above that is real signal. 

- `brier_class`: Mean squared error between predicted probabilities and the binary sepsis label 0/1 outcome. Measures **calibration + discrimination together** a model that says "80% sepsis risk" should be right 80% of the time. Lower is better; 0 is perfect, 0.25 is random guessing at 50/50. Unlike AUC metrics, Brier penalises *over-confident* probabilities

- `accuracy`: (TP+TN)/N. Due to imbalance this is included for familiarity and catching bugs! Reminder: Predicting "No Sepsis" for everyone scores ~98% here. 

- `sensitivity` (= recall, = true positive rate) TP/(TP+FN). Of patients who actually had sepsis, what fraction did we flag? The clinically critical metric for the positive class: missed sepsis (low sensitivity) means delayed antibiotics, which directly increases mortality.

- `specificity`: TN/(TN+FP). Of patients who did *not* have sepsis, what fraction did we correctly leave unflagged? The counterpart to sensitivity — high specificity means low false-alarm rate, which matters for **alarm fatigue** in the ICU. Sensitivity and specificity trade off as you move the threshold, so they're reported together.

Now we can run it, this will take a few minutes so we can use cache-ing again to avoid frustration!

```{r tune, cache = TRUE}
cls_metrics <- metric_set(pr_auc, brier_class, accuracy, sensitivity, specificity)

tuned <- wf_set |>
  workflow_map(
    fn        = "tune_grid",
    seed      = GLOBAL_SEED ,
    resamples = folds,
    grid      = 10,
    metrics   = cls_metrics,
    control   = control_grid(save_pred = TRUE, verbose = FALSE)
  )
```

### 5.5 Compare cross-validation results

```{r cv-comparison, fig.height = 3}
autoplot(tuned, metric = "pr_auc", select_best = TRUE) +
  labs(title = "Best CV PR-AUC per model family")
```

**Q13 (2 marks).** Which model family achieves the highest CV PR-AUC?
Is the gap between the best and worst model large relative to the standard errors?
What does that suggest about model selection here?

The logistic regression with the L1 and L2 correction achieved the highest PR-AUC (~0.76). The model with the worst PR-AUC of ~0.67 was the boost tree. The gap between these two models appears larger on the plot but is actually ~0.09 which is quite small. Moreover, their standard errors overlap which suggests that these differences in model performance are not statistically significant and model performance might actually be anywhere in that range. It is best to choose the highest performing model (logistic regression with L1 and L2 correction), however, any of these models would perform ~basically just as well (i.e., ~0.05 to ~0.09). 

```{r}
rank_results(tuned, rank_metric = "pr_auc", select_best = TRUE) |>
  filter(.metric %in% c("pr_auc", "brier_class", "accuracy", "sensitivity", "specificity")) |>
  select(wflow_id, .metric, mean, std_err, rank)
```

**Q14 (3 marks).** By summarising and comparing the metrics output by this code, describe how each model performed relative to one another.

While the three model types, logistic regression (glm_glm), random forest (rf_rf), and boost tree (xgb_xgb), are ranked 1-3 in the output, the values for each metric determining these ranks are only nominally different (i.e., in most cases to the second decimal point). We can ignore accuracy as all models perform the same at 98% (as noted above). In terms of brier class, all models are nearly perfect with the logistic regression only performing ~0.003 better than the random forest and boost tree models. As for pr-auc, all models are performing well above the 0.02 baseline and are all approaching a perfect pr-auc score is 1.0 -- the difference between the highest peforming pr-auc of 0.76 and the worst at 0.67 is 0.09 which again, is only a minor difference. Again, all models are performing almost the same in terms of specificity at ~100%, which means there's nearly 0 false alarms. However, in terms of the models ability to correctly identify true sepsis positive cases, all models perform worse than chance (worse than 50%). Again, all models are performing at roughly 40% which means 60% of true sepsis positive cases are not being correctly identified.   

---

## 6 Test-Set Evaluation

In the real-world we'd try to optimise these models further and/or try some different architectures but lets just take our best model and continue to test evaluation.

### 6.1 Finalise the best model

```{r finalise}
best_id <- rank_results(tuned, rank_metric = "pr_auc",
                        select_best = TRUE) |>
  filter(.metric == "pr_auc") |>
  slice_head(n = 1) |>
  pull(wflow_id)

final_wf <- extract_workflow(tuned, id = best_id) |>
  finalize_workflow(
    extract_workflow_set_result(tuned, id = best_id) |>
      select_best(metric = "pr_auc")
  )

split <- make_splits(train, test)
final_fit <- last_fit(final_wf, split = split, metrics = cls_metrics)
collect_metrics(final_fit)
```


```{r perf-curves, fig.width = 5.5, fig.height = 4}
preds <- collect_predictions(final_fit)

preds |>
  pr_curve(truth = SepsisLabel, .pred_Sepsis,
           event_level = "first") |>
  autoplot() + labs(title = "Precision–Recall curve (test set)")
```

### 6.3 Confusion matrix

**Q15 (1 marks).** How many times were true sepsis cases classified as non-sepsis in the test set? Is this more or less than the correct sepsis predictions?

18 people who are truly sepsis positive were wrongfully classified as NOT having sepsis. This is more than true sepsis positive cases that were correctly identified by the model (11 sepsis cases correctly classified). See below.

```{r conf-matrix}
preds |>
  conf_mat(truth = SepsisLabel, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(title = "Confusion matrix (threshold = 0.5)")
```

---

## 7 Misclassification Analysis

### 7.1 Subgroup performance

Now let's investigate whether the model work equally well/badly across patient groups?

**Code block broken/error message so using your posted HTML to answer questions below** 


**Q16 (2 marks).** Is there a subgroup where the model performs noticeably worse than the overall PR-AUC? 
Suggest one data-driven reason and one clinical reason that could explain the gap.

All subgroups perform much worse that the overall PR-AUC. Comparing subgroups to each other in the age and sex strata, they all perform about the same with difference at roughly the second or third decimal. A data-driven explanation is imbalances in sepsis cases between subgroups in each strata. A possible clinical explanation is that since we know that sepsis is a heterogenous condition, presentation of sepsis cases by subgroups may present very differently (e.g., male/female, middle age/seniors). 

### 7.2 Variable importance

```{r variable-importance, fig.height = 8}
parsnip_fit <- extract_fit_parsnip(final_fit)

importance_tbl <- if (grepl("^rf", best_id)) {
  imp <- parsnip_fit$fit$variable.importance
  tibble(variable = names(imp), importance = unname(imp))
} else if (grepl("^xgb", best_id)) {
  xgboost::xgb.importance(model = parsnip_fit$fit) |>
    as_tibble() |>
    transmute(variable = Feature, importance = Gain)
} else {
  broom::tidy(parsnip_fit) |>
    filter(term != "(Intercept)") |>
    transmute(variable = term, importance = abs(estimate))
}

importance_tbl |>
  slice_max(importance, n = 20) |>
  ggplot(aes(importance, fct_reorder(variable, importance))) +
  geom_col(fill = "steelblue") +
  labs(title = paste("Top variable importances: ", best_id),
       x = "Importance", y = NULL)
```

**Q17 (2 marks).** Compare the top 3 most important variables to the three predictors you identified in Q2.
Do they match? Write one sentence about any variable that surprised you.

The predictors I hypothesized as strong predictors were: diastolic blood pressure, heart rate, and temperature. From the plot above, the top three most important predictors were: lactate, MAP, and WBC. One predictor that ranked relatively high was Unit 1 which is the variable for	"Administrative identifier for ICU unit (MICU)". One reason this could be an important predictor is if many sepsis cases were coming from this particular unit.

---

## 8 Discussion

**Q18 (3 marks).** In 3–5 sentences, address both of the following:
- What is the single biggest limitation of this analysis (consider the label definition, the static sampling approach, generalisability to other hospitals, etc.)?
- Would you recommend deploying a model trained this way in a real ICU? If yes, under what monitoring conditions? If no, what would need to change first?

I think generalisability to other hospital populations is the biggest limitation. Data comes from only 2 hospitals in the US. Even if the datset included other US hospitals, generalisability outside the US would be limited given contextual factors like demographics, healthcare system, quality of care, incidence/burden of sepsis.

No, I would not recommend deploying this model in a real ICU. I don't have enought subject knowledge to really critique the methods ("how" the model was trained), but I think the low sensitivity (true positive rate) demonstrated by all 3 model types is concerning. In this scenario, the goal is to identify positive cases over negative cases. The case imbalance likely drove the sensitivity and specificity values from Q13. Prioritizing/optimizing sensitivity should be changed before ever deploying the model in a real world setting. To do so, using a dataset with a better balance in outcomes is recommended.

---

## Session Info

```{r session-info}
sessionInfo()
```

## Credits

This practical is adapted from earlier versions of the course's medical databases lab and updated to use the [PhysioNet/CinC Challenge 2019 sepsis dataset](https://physionet.org/content/challenge-2019/1.0.0/) (Reyna et al., *Crit Care Med* 2020) to use a more modern `tidymodels` workflow.
